Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-26 papers/paper/ms.tex
update ms

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is measured
  from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  (harboring \nplwgyroagenograzing\ known planets) for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of
  sub-gigayear systems can be attributed to the star formation rate in
  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \ratiosfr$\pm$\uncratiosfr\
  over the past 3\,Gyr.  We observe this trend both in known planet
  hosts and in the parent stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and its confirmation helps clarify the
  age distribution of the known transiting exoplanets.
  %
\end{abstract}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
    rotation sequences of open clusters}{seventh-order polynomials
    fitted to the slow rotation sequences of open clusters \citep[see Section~3.3 of][]{Bouma_2023}}.  The 2-D histogram of the
    stellar sample (left) includes only apparently single stars near
    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
    temperatures of 3800--6200\,K.  The planet-hosts (right) require
    the same stellar cuts, and include only the confirmed and
    candidate planets described in Section~\ref{subsec:plflags}.  The
    number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
    the relevant panels.  A simple estimate for the number of young
    planets discovered by Kepler follows by counting points below the
    \deleted{mean }rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citet{McQuillan_2014} (hereafter \citetalias{McQuillan_2014}); the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.95\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Reported ages of individual stars in benchmark open clusters.}
    Each point denotes a star's reported age, normalized by the age of
    a stellar ensemble in which the star is a candidate member.
    Cluster membership was evaluated without knowledge of rotation.
    Although some field interlopers may be present in the membership
    lists, outliers can be compared between different methods on a
    relative basis.  Each study was cross-matched against the same
    cluster list; certain methods report ages for more stars than
    others.  Horizontal scatter is added to visually clarify the
    statistical age uncertainties.  At $<$1\,Gyr, our ages are
    generally accurate, and their statistical uncertainties match
    their dispersion.
	}
	\label{fig:agescalecompone}
\end{figure*}

Stars that formed in the same birth cluster are the gold standard for
the astronomical age scale \citep{Soderblom_2010}.  Before Gaia, the
few open clusters known in the Kepler field had been cataloged by
\citet{1864RSPT..154....1H}.  Gaia has enabled the discovery of
new stellar ensembles that are more diffuse, but which nonetheless share a
common age based on isochrone, rotation, and lithium dating.
Specifically in the Kepler field,
NGC6811 (1\,Gyr; \citealt{Curtis_2019_ngc6811}) and NGC6819 (2.5\,Gyr; \citealt{Meibom_2015}) 
have been known for centuries,
while 
Theia~520 ($\approx$300\,Myr; \citealt{2019AJ....158..122K}), Melange-3 ($\approx$150\,Myr; \citealt{Barber_2022}), and Cep-Her ($\approx$40\,Myr; \citealt{Bouma_2022b} and Kerr et al.~2024 submitted) are more recent discoveries.

@@

	\vspace{-0.35cm}
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
    }
%
%	\vspace{-0.35cm}
%	\subfloat{
%		\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
%	}
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived age distributions of its stars (left) and planet
  hosts (middle).  The top row shows all stars with temperatures of
  3800--6200\,K for which we calculated rotational ages.  Opaque lines
  impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
  evolutionary state; transparent lines do not\added{ (see
  Section~\ref{subsec:flags})}.  The \replaced{bottom}{middle} row shows
  stars with temperatures of 4400--5400\,K, which \replaced{have more precise
  ages due to their fast spin-down}{often have more precise ages}.  The statistical uncertainty for a
  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
  row.  
  Finally, the right panel compares the rotation-derived ages
  against star formation histories derived using CMD fitting (dashed
  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
  \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

@@

\begin{figure}[!t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Star symbols denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  If one were to instead
    select for precise ages ($t/\sigma_{\rm t}>3$), this would hide
    $\approx$ten $<$0.5\,Gyr planets, and add $\approx$fifteen
    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
    hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the
    planets to similarly no have quality flags raised. 
    \label{fig:rp_period_age_results}
  }
\end{figure}

@@

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The SFH from
\citet{2019A&A...624L...1M} seems entirely consistent, particularly
after accounting for the statistical uncertainties of that study.
Regarding the episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a hint of the
$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
$\approx$1\,Gyr spike however is not recovered.  This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
\replaced{those studies is their improved accuracy}{previous work is that they are demonstrably accurate} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.

@@

{\it Binaries}:
\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
are biased toward rapid rotation \citep[e.g.][and many studies
thereafter]{Meibom_2007}.}{Stars in binaries are often biased
toward rapid rotation, even beyond the typical
realm of tidal synchronization \citep[see][]{Meibom_2007,Gruner_2023}.}
\added{\citet{2019ApJ...871..174S} for instance found that 59\% of
Kepler stars with rotation periods between 1.5 and 7\,days lie
0.3\,mag above the main sequence, compared with 28\% of their full
rotation sample.
We attempted to mitigate the effects of binarity with five distinct
quality flags (Section~\ref{subsec:flags}).   This attempt was at least
somewhat effective, based on the peak near zero in the raw
distributions of Figure~\ref{fig:hist_tgyro} being flattened in the
cleaned distributions.
However, the completeness of our quality flags is hard to assess.  One specific concern is that these
flags might have minimal sensitivity to binaries with
separations of $\approx$0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.  }

@@

Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
$P_{\rm rot}$$<$10\,day stars.
Relative to the nominal ``gyro applicable'' age distribution, there are three important differences.
There is a much larger drop in the star formation rate over the past three gigayears,
by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr.
The rate of decline in star formation
since $\approx$2.8\,Gyr ago is also constant, rather than having a ``kink'' at 1.5\,Gyr.
And finally, the corrected distributions show closer agreement with SFHs
derived from CMD fitting.
We caution however that this correction assumes that {\it none} of our
quality flags were able to identify any of these unresolved binaries.
Since we did in fact include flags based on e.g.,~location in the HR diagram,
this correction is best interpreted as an upper limit on the possible impact of binarity.}

@@

\added{{\it Weakened magnetic braking:} 
If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars could contaminate the peaks in
Figure~\ref{fig:hist_tgyro}.  This
scenario deserves consideration, given asteroseismic evidence for slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the observation of a ``long-period edge''
	in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
	may pile up over similar timescales \citep{David_2022}.  However, 
	this pileup is thought to happen at $>$5\,Gyr for
	late-G dwarfs and early K dwarfs \citep[see Figure~4
	by][]{2024ApJ...962..138S}.  This means that it cannot explain the
	2.3\,Gyr peak that we see in the 4400-5400\,K age distribution in Figure~\ref{fig:hist_tgyro}. 
	Independently,
	the analysis by \citet{2022ApJ...937...94M} fitted
	isochrones to a high-quality subset of Kepler data, and found that the
	\citet{McQuillan_2014} sample, discussed in Appendix~\ref{app:mcqonly}, does not
	include many stars older than $\approx$\,4\,Gyr.
These lines of evidence suggest that the peak
in the observed age distribution is not caused by a pile-up of
much older stars with especially lethargic spin-down rates, but that it could be
	contaminated by them beyond $\approx$3\,Gyr, especially for stars hotter than the Sun.
}

@@

The sizes and periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets
are mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an
evolutionary connection to their youth.  Other discoveries include
Earth-sized planets \added{such as Kepler-1312~c and Kepler-1561~b }with new ages of only a few hundred million years.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet-hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
    formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago\added{.  The statistical uncertainty on this figure is far smaller than the systematic uncertainty caused by the potential presence of unresolved photometric binaries (Section~\ref{subsec:caveats}); the star formation rate could have decreased by a factor of as much as 4.81$\pm$0.26}.  \replaced{This}{Either} result \deleted{from
    rotation-based ages }broadly agrees with recent reports of a
    declining star-formation rate from CMD fitting and white-dwarf
    chronology, though with the \added{potential }advantage of better sensitivity for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we should instead expect $\approx$250.
  %
  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

@@

\paragraph{Kepler-786}---This K3 V star has a surprisingly high
lithium content.  The spectrum is single-lined, with
\texttt{SpecMatch-Synth} derived parameters of $T_{\rm
eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1, [Fe/H]=0.11$\pm$0.06, and
$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm
Li}^{\ast}$=83$\pm$6\,m\AA, the lithium age of \kepseveneightsix\,Myr
would predict an obvious rotation signal.  None is present in the
Kepler light curve, consistent with the low $v\sin i$.  Out of all
``apparently discrepant'' lithium and rotation measurements discussed
in this appendix, this is the only one that seems to remain discrepant
after scrutiny.  The  \ion{Ca}{2} doublet is in emission, with
$R'_{\rm HK}=-4.7\pm0.5$, which suggests an age at least as old as the
Hyades \citep{Mamajek_2008}.  The Balmer lines are in absorption, and
display no obvious signatures of youth.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Variant of Figure~\ref{fig:prot_vs_teff}}, comparing the default
    rotation periods ({\it left}) from \citet{Santos_2019,Santos_2021}
    against the \citet{McQuillan_2014} rotation periods ({\it right}).
    All stars are required to be amenable to rotation-based age dating
    (i.e.~none of bit flags zero through \replaced{nine}{ten} raised).
	}
	\label{fig:prot_vs_teff_mcq}
\end{figure*}

>>>>

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is measured
  from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  (harboring \nplwgyroagenograzing\ known planets) for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of
  sub-gigayear systems can be attributed to the star formation rate in
  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{a few}
  over the past 3\,Gyr.  We observe this trend both in known planet
  hosts and in the parent stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and its confirmation helps clarify the
  age distribution of the known transiting exoplanets.
  %
\end{abstract}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.png}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
    rotation sequences of open clusters}{seventh-order polynomials
    fitted to the slow rotation sequences of open clusters \citep[see Section~3.3 of][]{Bouma_2023}}.  The 2-D histogram of the
    stellar sample (left) includes only apparently single stars near
    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
    temperatures of 3800--6200\,K.  The planet-hosts (right) require
    the same stellar cuts, and include only the confirmed and
    candidate planets described in Section~\ref{subsec:plflags}.  The
    number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
    the relevant panels.  A simple estimate for the number of young
    planets discovered by Kepler follows by counting points below the
    \deleted{mean }rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citet{McQuillan_2014} (hereafter \citetalias{McQuillan_2014}); the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(such as giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
signals have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.95\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Reported ages of individual stars in benchmark open clusters.}
    Each point denotes a star's reported age, normalized by the age of
    a stellar ensemble in which the star is a candidate member.
    Cluster membership was evaluated without knowledge of rotation\added{, and for this plot,
    no attempt was made to exclude known binaries}.
    Although some field interlopers may be present in the membership
    lists, outliers can be compared between different methods on a
    relative basis.  Each study was cross-matched against the same
    cluster list; certain methods report ages for more stars than
    others.  Horizontal scatter is added to visually clarify the
    statistical age uncertainties.  At $<$1\,Gyr, our ages are
    generally accurate, and their statistical uncertainties match
    their dispersion.
	}
	\label{fig:agescalecompone}
\end{figure*}

Stars that formed in the same birth cluster are the gold standard for
the astronomical age scale \citep{Soderblom_2010}.  Before Gaia, the
few open clusters known in the Kepler field had been cataloged by
\citet{1864RSPT..154....1H}.  Gaia has enabled the discovery of
new stellar ensembles that are more diffuse, but which nonetheless share a
common age based on isochrone, rotation, and lithium dating.
Specifically in the Kepler field,
NGC6811 (1\,Gyr; \citealt{Curtis_2019_ngc6811}) and NGC6819 (2.5\,Gyr; \citealt{Meibom_2015}) 
have been known for centuries,
while 
Theia~520 ($\approx$300\,Myr; \citealt{2019AJ....158..122K}), Melange-3 ($\approx$150\,Myr; \citealt{Barber_2022}), and Cep-Her ($\approx$40\,Myr; \citealt{Bouma_2022b} and \citealt{Kerr2024}) are more recent discoveries.

@@

	\vspace{-0.35cm}
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
    }
%
%	\vspace{-0.35cm}
%	\subfloat{
%		\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
%	}
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived age distributions of its stars (left) and planet
  hosts (middle).  The top row shows \replaced{all stars with temperatures of
  3800--6200\,K for which we calculated rotational ages}{stars with temperatures of 3800--6200\,K}.  Opaque lines
  impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
  evolutionary state; transparent lines do not\added{ (see
  Section~\ref{subsec:flags})}.  The bottom row shows
  stars with temperatures of 4400--5400\,K, which \replaced{have more precise
  ages due to their fast spin-down}{usually have more precise ages}.  The statistical uncertainty for a
  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
  row.  
  Finally, the right panel compares the rotation-derived ages
  against star formation histories derived using CMD fitting (dashed
  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr\added{, and decreases rapidly at older ages}
  \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

@@

\begin{figure}[!t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Star symbols denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  If one were to instead
    select for precise ages ($t/\sigma_{\rm t}>3$), this would hide
    $\approx$ten $<$0.5\,Gyr planets, and add $\approx$fifteen
    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
    hosts to not have $Q_{\rm star}$ bits 0--\replaced{9}{10} raised, and for the
    planets to similarly no have quality flags raised. 
    \label{fig:rp_period_age_results}
  }
\end{figure}

@@

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The SFH from
\citet{2019A&A...624L...1M} seems entirely consistent, particularly
after accounting for the statistical uncertainties of that study.
Regarding the episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a \added{weak }hint of the
\added{$\approx$1\,Gyr and }$\approx$2\,Gyr spike\added{s} reported by \citet{2020NatAs...4..965R}.  \deleted{The
$\approx$1\,Gyr spike however is not recovered. } This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
\replaced{those studies is their improved accuracy}{previous work is that they are demonstrably accurate} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.

@@

{\it Binaries}:
\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
are biased toward rapid rotation \citep[e.g.][and many studies
thereafter]{Meibom_2007}.}{Stars in binaries are often biased
toward rapid rotation, even beyond the typical
realm of tidal synchronization \citep[see][]{Meibom_2007,Gruner_2023}.}
\added{\citet{2019ApJ...871..174S} for instance found that 59\% of
Kepler stars with rotation periods between 1.5 and 7\,days lie
0.3\,mag above the main sequence, compared with 28\% of their full
rotation sample.
We attempted to mitigate the effects of binarity with five distinct
quality flags (Section~\ref{subsec:flags}).   This attempt was at least
somewhat effective, based on the peak near zero in the raw
distributions of Figure~\ref{fig:hist_tgyro} being flattened in the
cleaned distributions.
However, the completeness of our quality flags is hard to assess.  One specific concern is that these
flags might have minimal sensitivity to binaries with
separations of $\approx$0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.}

@@

Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
$P_{\rm rot}$$<$10\,day stars.
Relative to the nominal ``gyro applicable'' age distribution, there are three important differences.
There is a much larger drop in the star formation rate over the past three gigayears,
by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr.
The rate of decline in star formation
since $\approx$2.8\,Gyr ago is also constant, rather than having a ``kink'' at 1.5\,Gyr.
And finally, the corrected distributions show closer agreement with both the SFHs
derived from CMD fitting, and also with the KOI host distribution,
which is almost devoid of close binaries within 1\,AU \citep[e.g.][and references therein]{Moe2021}.
We caution however that this correction assumes that {\it none} of our
quality flags were able to identify any of these unresolved binaries.
Since we did in fact include flags based on e.g.,~location in the HR diagram,
this correction is best interpreted as an upper limit on the possible impact of binarity.}

@@

\added{{\it Weakened magnetic braking:} 
If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars could contaminate the peaks in
Figure~\ref{fig:hist_tgyro}.  This
scenario deserves consideration, given asteroseismic evidence for slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the observation of a ``long-period edge''
	in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
	may pile up over similar timescales \citep{David_2022}.  However, 
	this pileup is thought to happen at $>$5\,Gyr for
	late-G dwarfs and early K dwarfs \citep[see Figure~4
	by][]{2024ApJ...962..138S}.  This means that it cannot explain the
	2.3\,Gyr peak that we see in the 4400-5400\,K age distribution in Figure~\ref{fig:hist_tgyro}. 
	Independently,
	the analysis by \citet{2022ApJ...937...94M} fitted
	isochrones to a high-quality subset of Kepler data, and found that the
	\citet{McQuillan_2014} sample, discussed in Appendix~\ref{app:mcqonly}, does not
	include many stars older than $\approx$\,4\,Gyr.
These arguments suggest that the peak
in the age distribution is not caused by a pile-up of
much older stars with lethargic spin-down rates, but that it could be
	contaminated by them beyond $\approx$3\,Gyr, especially for stars hotter than the Sun.
}

@@

The sizes and periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets
are mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an
evolutionary connection to their youth.  Other discoveries include
Earth-sized planets \added{(e.g.~Kepler-1312~c and Kepler-1561~b) }with \deleted{new }ages of only a few hundred million years.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet-hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
    formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago\added{.  The statistical uncertainty on this figure is far smaller than the systematic uncertainty caused by the potential presence of unresolved photometric binaries (Section~\ref{subsec:caveats}); the star formation rate could have decreased by a factor of as much as 4.81$\pm$0.26}.  \replaced{This}{Either} result \replaced{from
    rotation-based ages broadly agrees}{is generally consistent} with recent reports of a
    declining star-formation rate from CMD fitting and white-dwarf
    chronology, though with the \added{potential }advantage of better \replaced{sensitivity}{accuracy} for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we should instead expect $\approx$250.
  %
  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

@@

\paragraph{Kepler-786}---This K3 V star has a surprisingly high
lithium content.  The spectrum is single-lined, with
\texttt{SpecMatch-Synth} derived parameters of $T_{\rm
eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1, [Fe/H]=0.11$\pm$0.06, and
$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm
Li}^{\ast}$=83$\pm$6\,m\AA, the lithium age of \kepseveneightsix\,Myr
would predict \replaced{an obvious}{a short-period} rotation signal.  \replaced{None is present in the
Kepler light curve}{The actual signal that is present has an amplitude of $\approx$0.1-0.3\% and a period of $\approx$33\,days}, consistent with the low $v\sin i$.  Out of all
``apparently discrepant'' lithium and rotation measurements discussed
in this appendix, this is the only one that seems to remain discrepant
after scrutiny.  The  \ion{Ca}{2} doublet is in emission, with
$R'_{\rm HK}=-4.7\pm0.5$, which suggests an age at least as old as the
Hyades \citep{Mamajek_2008}.  The Balmer lines are in absorption, and
display no obvious signatures of youth.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.44\textwidth]{prot_teff_Santos19_Santos21_dquality.png}
		}
		\subfloat{
			\includegraphics[width=0.44\textwidth]{prot_teff_McQuillan2014only_dquality.png}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Variant of Figure~\ref{fig:prot_vs_teff}}, comparing the default
    rotation periods ({\it left}) from \citet{Santos_2019,Santos_2021}
    against the \citet{McQuillan_2014} rotation periods ({\it right}).
    All stars are required to be amenable to rotation-based age dating
    (i.e.~none of bit flags zero through \replaced{nine}{ten} raised).
	}
	\label{fig:prot_vs_teff_mcq}
\end{figure*}
